Formal Analysis of EDHOC Key Establishment for Constrained IoT Devices

### 2022-07-12 protocol.tex
Moved content out into paper.tex and added intro.tex and protocol.tex to git

We denote by $\langle d_{t, \mathit{id}}, Q_{t, \mathit{id}}\rangle$ public-private key pairs, where $d$ is the private key, $Q$ is the public key, $t \in \{e, s\}$ indicates whether the key is ephemeral or static, and $\mathit{id}$ indicates the party to whom the key pair belongs.
%
When clear from context, we will often drop part (or all) of the subscript.
%
Static key pairs (suitable for both regular and challenge-response signatures) are long-term authentication credentials, while ephemeral key pairs are those which are generated afresh for each instantiation of the protocol.
%

Parties can authenticate using regular signatures or challenge-response signatures. 
%
In the former case, we say that they use the \emph{signature based authentication method} (\mSig{}).
%
In the latter, we say, following the terminology in the \mSpec{}, that they use the \emph{static key authentication method} (\mStat{}).
%
For more details about the challenge-response terminology, see~\cite{DBLP:conf/crypto/Krawczyk05}.

%%------------------------------------------------------------------------- sub
%

\mEdhoc{} relies on \mCose{}~\cite{rfc8152} for elliptic curve operations and
transforming points into bitstrings, and we therefore abstract those as
follows.
%
Signatures and verification thereof using party $A$'s key pair are
denoted \mSign{A}$(\cdot)$ and \mVf{A}$(\cdot)$ respectively.
%
The DH-primitive combining a private key $d$ and a point $P$ is denoted
$\mDH(d,P)$.
%
We abuse notation and let these function symbols denote operations on both
points and the corresponding bitstrings.
%

%------------------------------------------------------------------------- sub
\subsection{Framework Structure}
\label{sec:framework}
\mEdhoc{}'s goal is to establish an \mOscore{} security context,
including session key material \mbox{denoted \mSessKey{}}, and optionally transfer
application data \mADone{}, \mADtwo{} and \mADthree{}.
%
To accomplish this, the
\mSpec{}~\cite{our-analysis-selander-lake-edhoc-00} gives
a three-message protocol pattern, shown in Figure~\ref{fig:edhocFramework}.
%
We first describe this pattern and the parts that are common to all key
establishment methods.
%
Then we describe authentication and derivation of keys in more detail.
%
The latter is what differentiate the key establishment methods from each other.
%

\subsubsection{Protocol Pattern}
The first two messages negotiate authentication methods \mMethod{} and a
ciphersuite \mSuites{}.
%
In \mMethod{}, the initiator $I$ proposes which authentication method each
party should use.
%
These may differ, leading to four possible combinations:
\mSigSig{}, \mSigStat{}, \mStatSig{} and \mStatStat{}.
%
We refer to these \emph{combinations} of authentication methods simply as
\emph{methods} to align with the \mSpec{} terminology.
%
The first authentication method in a combination is the one proposed for the
initiator and the last is the one proposed for the responder.
%
The responder $R$ may reject the choice of method
or cipher suite with an error message, resulting in negotiation across
multiple \mEdhoc{} sessions.
%
Our analysis excludes error messages.
%

>>>>

%------------------------------------------------------------------------- sub
\subsection{Notation}
\label{sec:notation}
We denote by $\langle d_{t, \mathit{id}}, Q_{t, \mathit{id}}\rangle$ public-private key pairs, where $d$ is the private key, $Q$ is the public key, $t \in \{e, s\}$ indicates whether the key is ephemeral or static, and $\mathit{id}$ indicates the party to whom the key pair belongs.
%
When clear from context, we will often drop part (or all) of the subscript.
%
Static key pairs (suitable for both regular and challenge-response signatures) are long-term authentication credentials, while ephemeral key pairs are those which are generated afresh for each instantiation of the protocol.
%

Parties can authenticate using regular signatures or challenge-response signatures. 
%
In the former case, we say that they use the \emph{signature based authentication method} (\mSig{}).
%
In the latter, we say, following the terminology in the \mSpec{}, that they use the \emph{static key authentication method} (\mStat{}).
%
For more details about the challenge-response terminology, see~\cite{DBLP:conf/crypto/Krawczyk05}.
%

\mEdhoc{} fundamentally utilizes elliptic curves and associated Diffie-Hellman operations. 
%
Signatures using a principal $A$'s keys are denoted by \mSign{A}$(\cdot)$, while the verification thereof is denoted by \mVf{A}$(\cdot)$.
%
A Diffie-Hellman operation which combines a private key $d$ and a point $P$ on the elliptic curve is represented as $\mDH(d, P)$.
%
We will often overload notation to let $P$ stand for both the point on the elliptic curve as well as the corresponding bitstring.

%\mEdhoc{} relies on \mCose{}~\cite{rfc8152} for elliptic curve operations and
%transforming points into bitstrings, and we therefore abstract those as
%follows.
%
%Signatures and verification thereof using party $A$'s key pair are
%denoted \mSign{A}$(\cdot)$ and \mVf{A}$(\cdot)$ respectively.
%%
%The DH-primitive combining a private key $d$ and a point $P$ is denoted
%$\mDH(d,P)$.
%%
%We abuse notation and let these function symbols denote operations on both
%points and the corresponding bitstrings.
%

%------------------------------------------------------------------------- sub
\subsection{Overall description}
\label{sec:description}
\mEdhoc{} is designed to establish a security context for \mOscore{}.
%
This context includes the session key material (we denote this by \mSessKey{}), and might also transfer application data \mADone{}, \mADtwo{}, and \mADthree{}.
%
The abstract protocol as given in the \mSpec{}~\cite{} consists of three messages, and is shown in Figure~\ref{fig:edhocFramework}~\cite{Norr21}.
%
The generalized abstract protocol is the same across methods, but the authentication mechanisms and key derivation procedures differ.
%

Of the three messages, the first two, among other things, serve to establish a common authentication method \mMethod{}, and a ciphersuite \mSuites{}.
%
In \mMethod{}, the party playing the Initiator role proposes an ordered list of choices for the authentication methods that each party can use. 
%
As mentioned earlier, these may differ for the two roles, yielding four possible combinations: \mSigSig{}, \mSigStat{}, \mStatSig{}, and \mStatStat{}, where the first is the method used by the Initiator role, and the second by the Responder role.~\footnote{As in the \mSpec{}, we will call the combinations of authentication methods as methods as well.}
%
The party executing the Responder role may choose to reject the method or ciphersuite chosen by the Initiator by sending an error message.
%
This results in abandoning this session and renegotiating, as the Initiator goes down their list of choices for authentication methods, and picks the next option.
%
While our analysis (and indeed, \mTamarin{}) cannot naturally handle such renegotiation which maintains state and ``remembers'' the rejected method and/or ciphersuite, the formal modeling excludes this aspect of the protocol. 
%
However, we will discuss the ramifications of such a renegotiation procedure and the error messages later, in Section~\ref{sec:}.
%

In addition to negotiating the method and the ciphersuite, the first two messages are also instrumental for the exchange of public ephemeral keys \mGx{} and \mGy{}, and connection identifiers \mCi{} and \mCr{}, for the initiator and responder roles respectively. 
%
The \mSpec{} states that the connection identifiers serve only to route messages to the correct principal executing \mEdhoc{}, but then also claims that they may be used in turn by any protocols (like \mOscore{}) using the security context established by \mEdhoc{}. 
%
Thus, no explicit security guarantees are required to be satisfied by these communication identifiers, as per the \mSpec{}.

%%%%%%%% START HERE

%\mEdhoc{}'s goal is to establish an \mOscore{} security context,
%including session key material \mbox{denoted \mSessKey{}}, and optionally transfer
%application data \mADone{}, \mADtwo{} and \mADthree{}.
%%
%To accomplish this, the
%\mSpec{}~\cite{our-analysis-selander-lake-edhoc-00} gives
%a three-message protocol pattern, shown in Figure~\ref{fig:edhocFramework}.
%
%We first describe this pattern and the parts that are common to all key
%establishment methods.
%%
%Then we describe authentication and derivation of keys in more detail.
%%
%The latter is what differentiate the key establishment methods from each other.
%%
%
%\subsubsection{Protocol Pattern}
%The first two messages negotiate authentication methods \mMethod{} and a
%ciphersuite \mSuites{}.
%%
%In \mMethod{}, the initiator $I$ proposes which authentication method each
%party should use.
%%
%These may differ, leading to four possible combinations:
%\mSigSig{}, \mSigStat{}, \mStatSig{} and \mStatStat{}.
%%
%We refer to these \emph{combinations} of authentication methods simply as
%\emph{methods} to align with the \mSpec{} terminology.
%%
%The first authentication method in a combination is the one proposed for the
%initiator and the last is the one proposed for the responder.
%%
%The responder $R$ may reject the choice of method
%or cipher suite with an error message, resulting in negotiation across
%multiple \mEdhoc{} sessions.
%%
%Our analysis excludes error messages.
%
